The XAI Paradox: Systems that Perform Well for the Wrong Reasons

### 2018-08-30 Paper.tex
Details

A question that must be asked in XAI research is whether or not the explanation that a XAI system gives indeed makes sense in the domain. That this is not a trivial issue is in particular suggested by the phenomenon of adversarial examples in image classification, where input images are slightly altered using a perturbation \cite{yuan2017adversarial}. To the human eye, there is no apparent difference between the original input image and its altered version, while a machine learning system makes a completely different classification after the image is altered. Another example is the classic double-spiral problem \cite{Kaufmann1989Learning}, in which neural networks are tasked with discriminating between two interlocking spirals given an x and y coordinate. Rather than actually learning the mathematical functions that define the spirals, the networks usually create a type of ``patch-wise'' lookup table, that divides the image into a number sectors, each belonging to one of the two spirals. This suggests that the understanding that the systems have differs strongly from that of humans, despite the fact that the system usually makes the correct classifications. In other words, systems perform well, but apparently for the wrong reasons. In and of itself, this is still impressive, as the machine learning system is able to extrapolate structures from the data that we as humans are not able to perceive, and consequently achieve a high performance. However, explanations that are generated using XAI techniques may not make sense to the users, countering the goals of transparancy and explainability.

@@

The experiment with the symmetrical Boolean functions has confirmed that the neural network and the decision tree algorithm can learn the XOR condition and its generalized form, the parity function, if there is no interference from other variables. Compared to the M-out-of-N and exact value function, the parity function seems to be the most difficult function for the machine learning systems to learn. Both neural networks and decision trees, however, can learn them with sufficient training. This is supported by previous research that shows that neural networks have the potential of being universal approximators \cite{hornik1991approximation}. Once multiple conditions define the output of a dataset, however, the systems are unable to learn the individual conditions as well. The accuracies on the general test sets are still quite high for all possible interactions between the functions, but the systems do not learn the functions themselves as well. This supports the claim that a high classification accuracy is no guarantee that the right rationale is used in the classification process. The accuracies on the general test set and on each of the functions individually can be seen in Table \ref{tbl:interactions_DT} and Table \ref{tbl:interactions_NN}. It is interesting to see how systems can perform with an accuracy of almost 100\% on the general test set, yet fail dramatically on the test sets of the individual functions. This indicates that they do not actually learn these functions, but rather a different function that somehow accurately maps the input to the output. There must therefore be a confounding structure within the data that the systems learn instead of the original, intended structure. The systems could potentially learn the intended structure if additional constraints are added during the learning phase  \cite{schwab2012learn}. However, this would not prove that the systems themselves are able to internalize the intended structure, but rather display a form of intelligence by proxy.

>>>>

A question that must be asked in XAI research is whether or not the explanation that a XAI system gives indeed makes sense in the domain. That this is not a trivial issue is in particular suggested by the phenomenon of adversarial examples in image classification, where input images are slightly altered using a perturbation \cite{yuan2017adversarial}. To the human eye, there is no apparent difference between the original input image and its altered version, while a machine learning system makes a completely different classification after the image is altered. Another example is the classic double-spiral problem \cite{Kaufmann1989Learning}, in which neural networks are tasked with discriminating between two interlocking spirals given an x and y coordinate. Rather than actually learning the mathematical functions that define the spirals, the networks create a hierarchical structure of regions that divide the image into a number sectors, each belonging to one of the two spirals. This suggests that the understanding that the systems have differs strongly from that of humans, despite the fact that the system usually makes the correct classifications. In other words, systems perform well, but apparently for the wrong reasons. In and of itself, this is still impressive, as the machine learning system is able to extrapolate structures from the data that we as humans are not able to perceive, and consequently achieve a high performance. However, explanations that are generated using XAI techniques may not make sense to the users, countering the goals of transparancy and explainability.

@@

The experiment with the symmetrical Boolean functions has confirmed that the neural network and the decision tree algorithm can learn the XOR condition and its generalized form, the parity function, if there is no interference from other variables. Compared to the M-out-of-N and exact value function, the parity function seems to be the most difficult function for the machine learning systems to learn. Both neural networks and decision trees, however, can learn them with sufficient training (cf. the fact that neural networks are universal approximators \cite{hornik1991approximation}). 
Once multiple conditions define the output of a dataset, however, the systems are unable to learn the individual conditions as well. The accuracies on the general test sets are still quite high for all possible interactions between the functions, but the systems do not learn the functions themselves as well. This supports the claim that a high classification accuracy is no guarantee that the right rationale is used in the classification process. The accuracies on the general test set and on each of the functions individually can be seen in Table \ref{tbl:interactions_DT} and Table \ref{tbl:interactions_NN}. It is interesting to see how systems can perform with an accuracy of almost 100\% on the general test set, yet fail dramatically on the test sets of the individual functions. This indicates that they do not actually learn these functions, but rather a different function that somehow accurately maps the input to the output. There must therefore be a confounding structure within the data that the systems learn instead of the original, intended structure. The systems could potentially learn the intended structure if additional constraints are added during the learning phase  \cite{schwab2012learn}. However, this would not prove that the systems themselves are able to internalize the intended structure, but rather display a form of intelligence by proxy.
